Temporal ordering and registration of images in studies of developmental dynamics

### 2014-05-28 image_analysis_paper/pnas_draft2.tex
added edits to paper and si

\DeclareMathSizes{9}{8}{7}{7}

\DeclareMathOperator*{\argmin}{arg\,min}

@@

\significancetext{blahhhhh}

@@

\begin{abstract}
In studies of development, researchers are often presented with cross-sectional data, where each data point is a sample from a population fixed at a
%slightly
different developmental time.
%
The goal is then to temporally order the data to reconstruct the developmental dynamics.
%
Data points in the form of two-dimensional image must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform by hand.
%
We present an automated approach to registering {\it and} temporally ordering cross-sectional data sets of images.
%
The mathematical techniques (vector diffusion maps) are applicable to a wide variety of data sets and
require little {\it a priori} knowledge of the image features or the developmental dynamics.
%
%Furthermore, these techniques can separate multiple trajectories within the same data set.
%
Furthermore, the approach can detect multiple sources of variability
and thus help visualize multiple trajectories within a data set.
%
We demonstrate the utility of these methods using a collection of images from a study of {\it Drosophila} embryogenesis.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time for the same embryo \cite{roelens2013live, keller2013imaging}.
%
In a cross-sectional study, developmental dynamics must be reconstructed from multiple embryos, each of which contributes only a single snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
%
Both of these sampling schemes have their advantages and limitations, and both are extensively used by biologists.
%%%
%%%YGK sounds condescending
%%%
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%%%
%%%YGK  one or two refs ?
%%%
%
In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in large-scale
%%%
%%%YGK  large scale ? high throughput
%%%
physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point but also at a different geometric orientation.
%
Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
a certain time window to which a collection of embryos belongs, e.g., \cite{ng2012large, richardson2014emage, castro2009automatic}.
%to high accuracy.
%
Furthermore, each embryo appears at an arbitrary translation and rotation within the image frame (see {\it SI Appendix}).
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
%
We show how existing dimensionality reduction algorithms can automate, greatly accelerate {\it and combine} both of these tasks.

An example where
temporal ordering and registration of images can be done manually
because the number of images is small and the differences between them are visually apparent is shown
in Figure~\ref{fig:fish}: a caricature of fish development which illustrates the processes of growth and patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the position of head and fins.
%
In contrast to this simple illustration, real data poses nontrivial challenges for both registration and temporal ordering.
%
In general, the landmarks needed for registration (here the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
%
Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Fish, each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Fish, now rotationally registered and temporally ordered. For this caricature, the registration and ordering is easy to do ``by hand" because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
\end{figure}

@@

We show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
%
Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}, especially in the last decade. 
%
Here, we use a manifold learning algorithm \cite{singer2012vector} for high-dimensional imaging data. % which both factors out geometric symmetries to register the images and extracts low-dimensional structure in the imaging data set.
%
Such techniques have been used for data mining in
applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition; here they are adapted
for simultaneous registration and temporal ordering of imaging data from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
The third set consists of data collected from both wild type and mutant embryos, and joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
%
Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
%
%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
%
Our approach is based on vector diffusion maps \cite{singer2012vector}, a general technique developed for data sets which contain two types of sources of variability:
geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation that allows us to simultaneously register and temporally order our images.
%

Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
%
%%%YGK to imply which rotations we need to have discussed it above
%%
Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
%
We assume that, in this set of images, this main direction of variability is parameterized by the developmental time of each embryo, so taht uncovering this direction will allow us to recover the developmental dynamics.
%
Unlike the noise in the cryo-EM application, which is approximately Gaussian and arises from the measurement instrument, the noise in our experiments arises primarily from interembryo variability and does not have a known simple structure.

%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration is shown in Figure~\ref{subfig:synch1}, where each image is depicted as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).
%
In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical};
however, when the images are noisy, some pairwise alignments may be inaccurate.
%
Using the alignment angles between all pairs of vectors (or images), angular synchronization finds the set of rotation angles (one angle for each vector) most consistent with all pairwise measurements (see {\it SI Appendix}).
%
In this schematic, registration via angular synchronization is trivial as the pairwise measurements contain no noise.
%
However, the algorithm has been shown to successfully register sets of vectors (or images) even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
%

%
After removing the variability due to translations and rotations of the images, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the variability in the data.
%%%
%%%YGK feels like repeats
%%%
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique.
%
This algorithm uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%

The idea is illustrated in Figure~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a ``good'' parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
%
This idea of locality is denoted by the color intensity of the edges (denoting edge weight) in Figure~\ref{subfig:dmaps1}.
%
Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in Figure~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be much higher dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will rely on diffusion maps to automatically uncover a parameterization of our high-dimensional data (see {\it SI Appendix}).
%
Diffusion maps will also generalize from one-dimensional nonlinear curves to higher-dimensional manifolds.
%
Thus, when our experiments have additional sources of variability (such as mutations) in addition to developmental dynamics, diffusion maps can uncover both of these variability sources, and provide an informative organization of our data. %We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.

Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
%
Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find parameterizations which are globally informative.
%
Vector diffusion maps combines both steps into a single computation;
the algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see {\it SI Appendix}).
%
It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
%

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of data points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit SI Appendix}), so that close data points are connected by dark (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
\end{figure}

@@

As a first illustration of this approach to temporal ordering and registration of images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
Figure~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, the spatially uniform two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

Previous studies \cite{rushlow2012temporal} established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo (see {\it SI Appendix}).
%
The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour \cite{lim2013kinetics}.
%
This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
%
Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see Figure~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and intensity of the dpERK signal in the lateral regions of the embryo monotonically increases across the data set.
%

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: nuclei located under common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see {\it SI Appendix}).
%
Spearman's rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.

Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
%
An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes.
%
However, in other applications where the size variation is significant or the developmental dynamics are more subtle, it may be important to factor out size or {\it scaling} symmetries.
%%%
%%%YGK   for teh "additional components" - don't have in the same sentence - say it somewhere else.... INCLUDE IT (maybe in the mutants)

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because this data set in Figure~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time,
established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time.
%
However, developmental dynamics are often significantly more complex, and nonlinear techniques such as (vector) diffusion maps are required to extract meaningful structure in the data.
%

Figure~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
%
These images cover a thirty minute time interval and span late cellularization (the early images in this data set overlap with the late images in the previous set) through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs.
%
As before, each image is an optical cross-section of an embryo fixed at a different developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
%
Instead of Dl, which is difficult to visualize at this point, we stained for Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
%

%
During this developmental time span, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK (red in the figure) again appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
During this time window, germband elongation causes cells from the ventral side to move towards the poles of the embryo, and then wrap around to the dorsal side.
%
At the end of this process, cells which were originally on the ventral side have migrated to the dorsal side of the embryo, causing similar ``omega'' shaped patterns (formed by Twi and dpERK) to arise on the dorsal side (shown in the late images in Figure~\ref{subfig:averaged_data2}).

>>>>

%\DeclareMathSizes{9}{8}{7}{7}

%\DeclareMathOperator*{\argmin}{arg\,min}

\newcommand{\SI}[0]{{\it SI Materials and Methods}}
\newcommand{\fig}[0]{Fig.}

@@

\significancetext{Cross-sectional imaging studies are commonly used to study developmental dynamics in multiple organisms. In these studies, a representative developmental trajectory is reconstructed from snapshots of different embryos, each of which has been arrested at a different point of its development. One of the first steps in this reconstruction involves registration and temporal ordering of images. Here we show that both of these tasks can be automated and combined in a single step using recently developed advances in data mining. To illustrate this approach, we temporally order and register several datasets collected in cross-sectional studies of cell signaling in the early fruit fly embryo.}

@@

\begin{abstract}

Imaging studies provide unique insights into dynamics of developmental processes. 
%
A large number of imaging studies are based on a cross-sectional experimental design, where developmental progress is arrested in a population of embryos, each of which is at a different point along its developmental trajectory. 
%
The goal is then to temporally order the data to reconstruct a representative developmental dynamics from snapshots provided by the collection of embryos.
%
Images of different biological samples must first be registered before they can be temporally ordered.
%
When such data sets are large, noisy, and/or if the developmental changes are subtle, these tasks can be difficult to perform manually.
%
We present an automated approach to registering and temporally ordering data from such cross-sectional studies.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require {\it a priori} knowledge of image features or developmental dynamics.
%
%Furthermore, these techniques can separate multiple trajectories within the same data set.
%
We demonstrate the utility of these methods using a collection of images from a cross-sectional study of cell signaling during dorsoventral patterning of the {\it Drosophila} embryo.
\end{abstract}

@@

\dropcap{E}xperimental studies of developmental dynamics fall in two broadly defined categories: longitudinal and cross-sectional \cite{diggle2002analysis}.
%
In longitudinal studies, developmental progress is monitored over time in the same embryo \cite{roelens2013live, keller2013imaging}.
%
In a cross-sectional study, developmental dynamics is reconstructed from multiple embryos, each of which contributes only a snapshot of a chemical or morphological process along its developmental trajectory \cite{jaeger2004dynamic, fowlkes2008quantitative}.
%
Here we focus on cross-sectional studies, which have a time-honored history and still present the only option for most organisms.
%
In a typical cross-sectional study, a group of embryos is fixed using a procedure that arrests their development and stained with chemicals that visualize a handful of cellular processes.
%
Fixed embryos are then imaged using any given number of microscopy techniques.
%
Recent advances in physical manipulation and imaging of embryos produce rapidly increasing volumes of cross-sectional data, in which every embryo is observed at a different developmental time point and at a different geometric orientation.
%
Importantly, the ``age'' of any given embryo arrested in its development is not quantitatively known; typically what is known is
a certain time window to which a collection of embryos belongs \cite{ng2012large, richardson2014emage, castro2009automatic}.
%to high accuracy.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must be spatially aligned or {\em registered} to factor out the relevant geometric symmetries (e.g., translations and rotations), and then ordered in time.
%
We show how existing dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

Temporal ordering and registration of images can be done manually
when the number of images is small and the differences between them are visually apparent. 
%
\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the position of head and fins.
%
In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
%
In general, the landmarks needed for registration (here, the heads and fins) as well as the attributes which can be used to order the data (such as the body length) are not known {\it a priori}.
%
Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig1}
\caption{Caricature illustrating the tasks of image registration and temporal ordering. {\xfigtextfontit (A)} Images of ``samples'', each in a different orientation and a different stage of development. {\xfigtextfontit (B)} Registered and ordered samples. For this caricature, the registration and ordering is straightforward because the data set is small and the developmental changes are easy to recognize.}
%\label{fig:fish}
\customlabel{fig:fish}{1}
\customlabel{subfig:fish_unordered}{\ref{fig:fish}{\it A}}
\customlabel{subfig:fish_ordered}{\ref{fig:fish}{\it B}}
\end{figure}

@@

We show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
%
Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}. 
%
Here, we use a manifold learning algorithm \cite{singer2012vector} that has been recently developed for high-dimensional imaging data. 
%
This technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of imaging studies. 
%
Here it is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

%%%YGK  when I read this I think of Eiffel tower pictures, not of ours - there is a transition... careful where we do it.
%
Our goal is to organize data sets from cross-sectional studies of developmental processes in a meaningful and informative way.
%
%We will show that dimensionality reduction algorithms can simultaneously register and temporally order the images to give us a meaningful picture of the underlying developmental dynamics.
%
The approach is based on vector diffusion maps \cite{singer2012vector}, a manifold learning
technique developed for data sets which contain two types of sources of variability:
geometric symmetries (such as translations and rotations of the images due to the experimental setup) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one wants to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \cite{singer2011angular} for image registration and {\em diffusion maps} \cite{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation that allows us to simultaneously register and temporally order our images.
%

Vector diffusion maps has been used to classify images from cryo-electron microscopy (cryo-EM) experiments, where each image is a projection of a randomly oriented molecule.
%
%%%YGK to imply which rotations we need to have discussed it above
%%
Here, we will use the algorithm to register images of {\it Drosophila} embryos with respect to rotations and translations, as well as uncover the main direction of variability {\it after} removing translational and rotational symmetries.
%
We assume that, in these sets of images, the main direction of variability is parameterized by the developmental time of each embryo, so that uncovering this direction will allow us to recover the developmental dynamics.

%%%somewhere I want a sentence that says this is for general symmetry groups of some kind, and HERE it is for..... (maybe in SI)
Angular synchronization uses pairwise alignment information to register a set of images in a globally consistent way.
%
A schematic illustration of angular synchronization is shown in \fig~\ref{subfig:synch1}, where each image is represented as a vector, and the goal is to align the set of vectors.
%
We first compute the angles needed to align pairs of vectors (or images).  
%
In general, this requires no template function \cite{ahuja2007template} or image landmarks \cite{ian1998statistical};
however, when the images are noisy, some pairwise alignments may be inaccurate.
%
Using the alignment angles between all pairs of vectors, angular synchronization finds the set of rotation angles (one angle for each vector) that is most consistent with all pairwise measurements (see {\it SI Materials and Methods}).
%
In this schematic, registration via angular synchronization is trivial as the pairwise measurements contain no noise.
%
However, the algorithm can successfully register data sets even when many of the pairwise measurements are corrupted by noise \cite{singer2011angular}.
%

%
After removing variability due to translations and rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional manifold that parameterizes most of the remaining variability in the data.
%%%
%%%YGK feels like repeats
%%%
%
Such a manifold can be discovered using diffusion maps \cite{coifman2005geometric}, a nonlinear dimensionality reduction technique that uncovers an intrinsic parameterization of data that lies on a low-dimensional, perhaps nonlinear, manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{subfig:dmaps1}, where the data are two-dimensional points which lie on a one-dimensional nonlinear curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry: we want points which are close in high-dimensional space (e.g., images which look similar) to be close in our parameterization.
%
This idea of locality is denoted by the edges in \fig~\ref{subfig:dmaps1}.
%
Data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{subfig:dmaps2} depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
In our working examples, each data point will be much higher dimensional (e.g, an image with many pixels), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps to automatically uncover a parameterization of our high-dimensional data (see \SI).
%
Diffusion maps also generalizes from one-dimensional nonlinear curves to higher-dimensional manifolds.
%
Thus, when our experiments have additional sources of variability (such as distinct populations of embryos) in addition to developmental dynamics, diffusion maps can provide an informative organization of our data which highlights both of these variability sources.
%We will assume that our data are approximately one-dimensional, and that this dimension is parameterized by time, such that ordering our data along this main dimension/direction will temporally order our data.

Both angular synchronization and diffusion maps use pairwise information to uncover global structure in the data.
%
Angular synchronization uses pairwise alignment information to find translations and rotations which are globally consistent, and diffusion maps uses pairwise distances to find parameterizations which are globally informative.
%
Vector diffusion maps combines both steps into a single computation.
%
The algorithm uses information about pairwise alignments {\em and} pairwise distances to find a global parameterization of the data, while registering images which are ``close'' (see \SI).
%
It is thus perfectly suited for our task of registering and temporally ordering cross-sectional imaging data.
%

\begin{figure}[t]
\includegraphics[width=8.7cm]{fig2}
\caption{Schematic illustrating angular synchronization and diffusion maps. {\xfigtextfontit (A)} Set of vectors, each in a different orientation. The pairwise alignment angles are indicated. {\xfigtextfontit (B)} The vectors from {\xfigtextfontit A}, each rotated about their midpoint so that the set is globally aligned. Note that the chosen rotation angles are consistent with the pairwise alignments in {\xfigtextfontit A}: the differences between a pair of angles in {\xfigtextfontit B} is the same as the pairwise angle in {\xfigtextfontit A}. {\xfigtextfontit (C)} Data points (in black) which lie on a one-dimensional nonlinear curve in two dimensions. Each pair of points is connected by an edge, and the edge weight is related to the Euclidean distance between the points through the diffusion kernel (see {\xfigtextfontit \SI}), so that close data points are connected by dark (``stronger'') edges. {\xfigtextfontit (D)} The data in {\xfigtextfontit C}, colored by the first (non-trivial) eigenvector from the diffusion map computational procedure. The color intensity is monotonic with the arclength, thus parameterizing the curve.}
%\label{fig:schematics}
\customlabel{fig:schematics}{2}
\customlabel{subfig:synch1}{\ref{fig:schematics}{\it A}}
\customlabel{subfig:synch2}{\ref{fig:schematics}{\it B}}
\customlabel{subfig:dmaps1}{\ref{fig:schematics}{\it C}}
\customlabel{subfig:dmaps2}{\ref{fig:schematics}{\it D}}
\end{figure}

@@

As a first illustration of using vector diffusion maps to temporally order and register images, we analyze a data set where the correct registration and temporal order is known {\it a priori}.
%
This data set is obtained by fluorescent imaging of {\it Drosophila} embryos during the third hour after fertilization.
%
By this stage of development, sequential divisions of the fertilized egg have generated a system where most of the nuclei are arranged in a monolayer under the common plasma membrane \cite{foe1983studies}.
%
\fig~\ref{subfig:raw_data1} shows optical cross-sections of each embryo, with nuclei (gray in the figure) labeled by DAPI, a DNA stain.
%
During the third hour of development, the spatially uniform two-dimensional arrangement of nuclei is patterned by the concentration profiles of chemical signals that control cell differentiation.
%
One of these signals is provided by the nuclear localization of the transcription factor Dorsal (Dl, green in the figure), which subdivides the early embryo into the territories that give rise to the muscle, nerve, and skin tissues.
%
Another signal is provided by the phosphorylated form of the extracellular signal regulated kinase (dpERK, red in the figure), an enzyme that specifies a subset of neuronal cells.
%

@@

Previous studies established that throughout the third hour of development, the nuclear localization of Dl is unimodal and peaked at the ventral side of the embryo, reviewed in \cite{rushlow2012temporal}.
%
The spatiotemporal pattern of dpERK is more complex: dpERK is first detected in two lateral stripes, which increase in intensity towards the end of the third hour of development \cite{lim2013kinetics}.
%
This bimodal pattern is then augmented by a peak at the dorsal side of the embryo.
%
Both of these features become readily apparent when images are registered and ordered using vector diffusion maps (see \fig~\ref{subfig:ordered_data1}): the peak of nuclear Dl is consistent throughout the images and the intensity of the dpERK signal in the lateral regions of the embryo increases monotonically across the data set.
%

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S1).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.

%Although the first vector diffusion maps coordinate accurately captures the developmental dynamics, there are other sources of variability within this data set that could also be extracted using these techniques.
%
%An obvious one is variability due to size; here, the size variation is relatively small relative to the developmental changes and does not significantly obstruct the reconstruction of the temporal order.
%
%However, in other applications where the size variation is significant %or the developmental dynamics are more subtle, it may be important to %factor out size or {\it scaling} symmetries.
%%%
%%%YGK   for teh "additional components" - don't have in the same sentence - say it somewhere else.... INCLUDE IT (maybe in the mutants)

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S6).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
%

\fig~\ref{subfig:raw_data2} shows a richer data set of $108$ images of {\it Drosophila} embryos.
%
These images cover a thirty minute time interval and span late cellularization through gastrulation, when the two-dimensional sheet of newly formed cells begins to deform into three-dimensional structures which will form the future organs \cite{leptin2005gastrulation}.
%
As before, each image is an optical cross-section of a vertically oriented embryo fixed at a different developmental time.
%
The nuclei (gray) are again labeled with a DNA marker.
%
However, during this developmental time period, the nuclei are no longer uniformly arranged around the periphery of the embryo.
%
Instead of Dl, which is difficult to visualize at this point, embryos were stained with the antibody that recognizes Twist (Twi, shown in green), a transcription factor that is activated by high levels of Dl, specifying the cells of the future muscle tissue.
%
dpERK is again shown in red.

%
During this window, the ventral furrow is formed, where the ventral side of the embryo buckles towards the center of the embryo, bringing in the future muscle cells.
%
dpERK appears as two lateral peaks at the ventral side of the embryo.
%
These two peaks merge together during invagination, eventually forming (together with Twi) a characteristic ``omega'' shape.
%
During this time window, germband extension causes cells from the ventral side to move towards the posterior pole of the embryo, and then wrap around to the dorsal side.
%
At the end of this process, cells which were originally on the ventral and posterior side of the embryo find themselves on the dorsal side, causing similar ``omega'' shaped patterns to arise on the dorsal side; these patterns are most readily seen in the last image of \fig~\ref{subfig:averaged_data2}.

\begin{figure*}[t]
\includegraphics{fig4}
\caption{{\xfigtextfontit Drosophila} during gastrulation. {\xfigtextfontit (A)} Images of gastrulating {\xfigtextfontit Drosophila} embryos. Each image is of a different embryo in a different rotational and translational orientation. {\xfigtextfontit (B)} Data from {\xfigtextfontit A}, registered and ordered using vector diffusion maps. {\xfigtextfontit (C)} Average trajectory of the images shown in {\xfigtextfontit B}. Each image is a (Gaussian-weighted) average of the images in {\xfigtextfontit B}. }
\customlabel{fig:data2}{4}
\customlabel{subfig:raw_data2}{\ref{fig:data2}{\it A}}
\customlabel{subfig:ordered_data2}{\ref{fig:data2}{\it B}}
\customlabel{subfig:averaged_data2}{\ref{fig:data2}{\it C}}
\end{figure*}
